---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ActDAnalyser

<!-- badges: start -->
<!-- badges: end -->

## Description

This package provides an easy way to analyse ActinomycinD RNA sequencing data
and to infer RNA half-life from it.

If you only want to access measured half-life you can use the [ActDBrowser](https://www.signalingsystems.ucla.edu/ActDBrowser), that 
will allow to search available datasets for measured half-life for specific genes or overall distributions. 

## Installation

You can install this package by using devtools package:    
```{r, eval=FALSE}
devtools::install_github("dlefaudeux-ucla/ActDAnalyser")
```
If you don't have devtools installed you can install it using:
```{r, eval=FALSE}
install.packages('devtools')
```

Once ActDAnalyser is installed then you can load it:
```{r}
library('ActDAnalyser')
```

## Tutorial

### Test Datasets
You can use the demo data BMDM_dataset or MEF_dataset. To get a description of 
the datatsets run for example:
```{r, eval=FALSE}
?BMDM_dataset
```

Let's work on the BMDM dataset. Attach the dataset to make all object available:
```{r}
attach(BMDM_dataset)
```

Thefollowing objects corresponds to object you would need to have to be able to 
run the pipeline:

* **full_count:** corresponds to the count tables, every column corresponds to one 
sample an every row to one gene. Moreover the row names must correspond to the 
gene id.
```{r}
full_counts[1:10,1:3]
str(full_counts)
```
* **gene_infos:** corresponds to additional information on the genes. It must 
contain a column named Geneid that match the row names of full_counts.
```{r}
str(gene_infos)
```
* **metadata:** corresponds to additional data related to the full_counts object. 
The metadata data.frame must have specific format. The column names must match 
exactly the ones in this example.     
The following columns are absolutely required:  
  * _ExperimentNumber:_ integer, must be identical for a single ActinomycinD timeserie
  * _ExperimentName:_ character, names given to single ActinomycinD timeserie. 
  It must correspond to a single ExperimentNumber
  * _ExperimentReplicate:_ integer, for each experiment a number to match specific 
  experiment replicate
  * _ActDTime:_ numeric, time between ActinomycinD treatment and sample sequencing. 
  Must be in minutes
  * _StimulusTime:_ numeric, for timeseries half-life estimation, time between stimulus
  and ActD treatment. Must be in minutes. I the experiments corresponds to steady-
  state half-life then set all the values in this column to 0
  * _SampleName:_ character, name of each sample. Must match the column names of the 
  full_counts
  * _QC:_ character, if for any a specific sample needs to be removed from the analysis
  set its value to 'QC failed' (it must be exactly written this way). Otherwise set
  all values to 'Included'
```{r}
str(metadata)
```

You can check your metadata dat.frame for obvious inconsistency wiyh:
```{r}
checkMetadata(metadata, full_counts, type = 'all')
```

### Prepocessing
First we remove genes that have low counts for all samples:
```{r}
filtered <- applyCountThreshold(data = full_counts, metadata = metadata, threshold=32)
```

Then we are going to separate spike-ins from the rest of the genes.
```{r}
tmp <- extractSpikeIn(data = filtered, spikein = "ERCC")
spikeins <- tmp[[1]]
counts <- tmp[[2]]
```

Then we can normalise the data based on the spike-in information
```{r,fig.width=10, fig.height=5}
tmp <- SpikeInsNormalisation(spikeins, counts, gene_infos, metadata)
norm_spikeins <- tmp[[1]]
norm_counts <-  tmp[[2]]
```

Now we can compare the raw data and the normalised data.
```{r plot_spikeins, fig.width=10, fig.height=5}
# Plot spike-ins counts before and after normalisation
plotSpikeIns(before = spikeins, after = norm_spikeins, metadata = metadata)
```

As we can see after normalisation the spike in counts are brought to similar levels.

```{r plot_library_size, fig.width=10, fig.height=5}
# Plot library size before and after normalisation
plotLibrarySize(before = counts, after = norm_counts, metadata = metadata)
```

As we can see most experiment show a decreasing library size after normalising, 
which is what we would expect because there will be less and less RNA given that
it is continuously degraded and non produced.    
Experiments that don't follow that pattern might be of lesser quality (i.e. the 
spike-ins addition to each library might be to variable to lead to accurate results) 
and the results might not be accurate

```{r plot_pca, fig.width=10, fig.height=5}
# plot pca of counts before/after normalisation
plotPCA(before = counts, after = norm_counts, metadata = metadata)
```

Simirlaly we can see that some sample behave weird and that is why they have been 
set as 'QC failed'

### Infering Half-life
To infer hal-life use this:
```{r}
decay <- inferDecay(norm_counts = norm_counts[1:1000,], metadata = metadata, threshold = 32)
halflife <- convertSlopeToHL(decay)
```

The decay object is used to visulaize specific genes.   

The half-life object recapitulate the main result. It is a list with one element 
per ActinomycinD experiment.

```{r}
head(halflife[[1]])
```

#### Visulasizing result
You can plot fit for specific genes using

```{r plot_gene_ensembl_id, fig.width=7, fig.height=5}
plotGeneFit(gene = "ENSMUSG00000033845.11", norm_counts = norm_counts, metadata = metadata,
            decay_result = decay, gene_infos = gene_infos, gene_type='Geneid', 
            control=list(facet_grid=list(rows = formula('~Conditioning')), theme = list(aspect.ratio=1)))
```

Given that the gene_infos data.frame also has Symbol we can also use the gene name to plot,
for that we need to specify to look in the Symbol column by setting `r gene_type = 'Symbol'`
```{r plot_gene_name, fig.width=7, fig.height=5}
plotGeneFit(gene = "Cd28", norm_counts = norm_counts, metadata = metadata,
            decay_result = decay, gene_infos = gene_infos, gene_type='Symbol', 
            control=list(facet_grid=list(rows = formula('~Conditioning')), theme = list(aspect.ratio=1)))
```

Additionally you can also plot the distribution of all the genes for a specific R² threshold.

```{r plot_distribution, fig.width=7, fig.height=5}
plotDistribution(HL = halflife, metadata = metadata ,
            control=list(theme = list(aspect.ratio=0.5)))
```

You can also only plot the distribution of genes having a adjusted R² within a specified interval. Here set to show distribution for fits having and adjusted R² between 0.99 and 1.

```{r plot_distribution_subset, fig.width=7, fig.height=5}
plotDistribution(HL = halflife, metadata = metadata , r2_th = c(0.99,1),
            control=list(theme = list(aspect.ratio=0.5)))
```
